---
title: "Pharmaceutical Report"
author: "Group 22"
format: typst
editor: visual
---

```{r setup}
#| include: false

# ---------------------------------------------------------------------------------
# Setup
# ---------------------------------------------------------------------------------

# Clear the environment
rm(list = ls())

# load the packages
pacman::p_load(
  tidyverse,           # Data wrangling + graphics
  tibble,              # Alternative to dataframes with slightly altered rules
  janitor,             # Tidy dataframe names
  scales,              # Graphics formatting
  MASS,                # Negative binomial models
  glmmTMB,             # Negative binomial mixed models
  pander               # clean tables
)

# Set the seed
set.seed(111)

# Load the data
Data_ByChemical <- read.csv("Data/Data_ByChemical.csv") |>
  as_tibble()
Deprivation <- read.csv("Data/Deprivation index 2023.csv") |>
  as_tibble() |>
  clean_names()
Retired_population <- read.csv("Data/Retired population annual estimates.csv") |>
  as_tibble() |>
  clean_names()


# Define deprivation order
deprivation_order <- c(
  "1 - Least deprived",
  "2", "3", "4", "5", "6", "7", "8", "9",
  "10 - Most deprived",
  "Not elsewhere included",
  "Total stated - New Zealand index of socioeconomic deprivation",
  "Total - New Zealand index of socioeconomic deprivation"
)

# Approximate geographic order from north to south
district_order <- c(
  "Northland",
  "Waitematā",
  "Auckland",
  "Counties Manukau",
  "Waikato",
  "Bay of Plenty",
  "Lakes",
  "Tairāwhiti",
  "Taranaki",
  "Whanganui",
  "MidCentral",
  "Hawke's Bay",
  "Capital, Coast and Hutt Valley",
  "Wairarapa",
  "Nelson Marlborough",
  "West Coast",
  "Canterbury",
  "South Canterbury",
  "Southern",
  "Area outside health region"
)

# Create a vector just for dementia medication
dementia_chemicals <- c(
  "Donepezil hydrochloride",
  "Rivastigmine"
)

# ---------------------------------------------------------------------------------
# Summaries
# ---------------------------------------------------------------------------------

# Remove the totals
Data_ByChemical_clean <- Data_ByChemical |>
  filter(District != "New Zealand") |>
  # Rename the districts for merging purposes
  mutate(
    District = case_when(
      District == "Tairawhiti" ~ "Tairāwhiti",
      District == "Waitemata" ~ "Waitematā",
      District %in% c("Capital & Coast", "Hutt Valley") ~ "Capital, Coast and Hutt Valley",
      .default = District
    ),
    NumDisps = as.numeric(
      case_when(
        NumDisps == "<6" ~ "3", 
        .default = NumDisps)
    ),
    NumPpl = as.numeric(
      case_when(
        NumPpl == "<6" ~ "3", 
        .default = NumPpl)
    )
  ) |>
  # Summarise the Capital, Coast and Hutt Valley instances
  group_by(Type, ChemID, Chemical, District, YearDisp, NHIComp) |>
  summarise(
    NumDisps = sum(NumDisps),
    NumPpl = sum(NumPpl)
  )

# ------------------------------------------------------------------------------
# Summaries
# ------------------------------------------------------------------------------

# 15 most common medicines by number of distributions
Common_chem_15 <- Data_ByChemical_clean |>
  group_by(Chemical) |>
  summarise(NumDisps = sum(NumDisps)) |>
  ungroup() |>
  slice_max(NumDisps, n = 15) |>
  pull(Chemical)

# - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
# Chemical dispension count
# - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 

Data_ByChemical_count <- Data_ByChemical_clean |>
  group_by(Chemical) |>
  summarise(NumDisps = sum(NumDisps)) |>
  ungroup() |>
  arrange(desc(NumDisps))

Data_ByChemical_year_count <- Data_ByChemical_clean |>
  group_by(YearDisp, Chemical) |>
  summarise(NumDisps = sum(NumDisps)) |>
  ungroup() |>
  arrange(YearDisp, desc(NumDisps)) |>
  group_by(YearDisp) |>
  mutate(Rank = row_number()) |>
  slice_min(Rank, n = 15) |>
  ungroup()

# - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
# Concentrated Medication by District
# - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 

# Find the regions where each medication has the highest concentration of
# distribution
Chemical_concentration_region <- Data_ByChemical_clean |>
  filter(District != "Unknown") |>
  group_by(Chemical, District) |>
  summarise(NumDisps = sum(NumDisps)) |>
  ungroup() |>
  group_by(Chemical) |>
  mutate(
    Chemical_total = sum(NumDisps),
    district_share = NumDisps / Chemical_total,
    districts_observed = n()
  ) |>
  slice_max(
    district_share, n = 1
  ) |>
  ungroup() |>
  arrange(desc(district_share))

# - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
# Deprivation pivot
# - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 

# Change the raw data to wide format
Deprivation_wide <- Deprivation |>
  dplyr::select(area, new_zealand_index_of_socioeconomic_deprivation, obs_value) |>
  pivot_wider(names_from = area, values_from = obs_value) |>
  mutate(
    new_zealand_index_of_socioeconomic_deprivation = factor(
      new_zealand_index_of_socioeconomic_deprivation,
      levels = deprivation_order,
      ordered = TRUE
    )
  ) |>
  arrange(new_zealand_index_of_socioeconomic_deprivation)

# Create a lookup of average deprivations per district
Deprivation_average <- Deprivation_wide |>
  mutate(
    index = parse_number(
      as.character(
        new_zealand_index_of_socioeconomic_deprivation
      )
    )
  ) |>
  filter(index %in% 1:10) |>
  pivot_longer(
    cols = -c(
      new_zealand_index_of_socioeconomic_deprivation,
      index
    ),
    names_to = "District",
    values_to = "population"
  ) |>
  group_by(District) |>
  summarise(
    average_deprivation = weighted.mean(
      index,
      population,
      na.rm = TRUE
    ),
    .groups = "drop"
  )

# - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
# Dementia medication
# - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 

dementia_data <- Data_ByChemical_clean |>
  filter(
    Chemical %in% dementia_chemicals,
    District != "Unknown"
  ) |>
  mutate(
    NumDisps = case_when(
      NumDisps == "<6" ~ 3,
      TRUE ~ as.numeric(NumDisps)
    ),
    NumPpl = case_when(
      NumPpl == "<6" ~ 3,
      TRUE ~ as.numeric(NumPpl)
    )
  ) |>
  group_by(YearDisp, District, Chemical) |>
  summarise(
    NumDisps = sum(NumDisps),
    NumPpl = sum(NumPpl),
    .groups = "drop"
  ) |>
  left_join(
    Deprivation_average,
    by = "District"
  )

# - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
# Population counts
# - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 

# Tidy the data
Retired_population_clean <- Retired_population |>
  filter(
    age != "65 years and over",
    area != "Area outside health district"
  ) |>
  mutate(
    YearDisp = year_at_30_june,
    age_group = age,
    age_lower = parse_number(age),
    population = obs_value,
    District = area
  ) |>
  dplyr::select(YearDisp, YearDisp, age_group, age_lower, District, population)

# Find the district populations for people aged 65 or over
population_65_plus <- Retired_population_clean |>
  group_by(
    YearDisp,
    District
  ) |>
  summarise(
    population_65_plus = sum(
      population
    )
  ) |>
  ungroup()

# Combine population and dispensing data
dementia_population <- dementia_data |>
  left_join(
    population_65_plus,
    by = c(
      "YearDisp",
      "District"
    )
  ) |>
  mutate(
    dispensings_per_1000_65plus =
      NumDisps / population_65_plus * 1000,
    
    people_per_1000_65plus =
      NumPpl / population_65_plus * 1000
  )
```

## Executive Summary

## Table of Contents

## Background and Data

[What is dementia and the importance of pharmaceutical treatement]{.underline}

[Importance of equity across socioeconomic status]{.underline}

[Pharmaceutical collection background]{.underline}

[Note that dispensing doesn't immediately imply that the medication was taken]{.underline}

[Association of dispensing between deprivation and time]{.underline}

[Background on the NZDep2023 population counts by deprivation decile and health disrtict, and the attributes included + supression rules and unknown districts]{.underline}

[The pharmaceutical data is a repeated cross sectional data frame showing how medicine was distributed in different regions over different years.]{.underline}

[Note that just because medicine was dispensed does not mean that it was taken.]{.underline}

[Additional district level variables to look into:\
Income, age, employment, education, racial profile (Probably not), family dependencies.]{.underline}

## Ethics, Privacy and Security

- [Ethical considerations:]{.underline}

  - [Any comparison of groups being done in this project shouldn't cause have any implications about financial status between said groups. For example, Māori people take less medication than Pākehā because they can't afford them.]{.underline}

  - [Given the nature of the data collection, no causal causal conclusions can be drawn. The analysis will only be able to create commentary about the trends and relationships between the variables.]{.underline}

- [Privacy considerations:]{.underline}

  - [We should not be removing the suppression layer from the count of medication distribution. While there is an oversight in how this data is provided that allows users to do this, and there is code demonstrating that this is possible in this project, the values were randomly adjusted so that for the example instance doesn't have the true count revealed.]{.underline}

  - [For all analysis, the suppressed values will be imputed. Currently the method of replacing these with the median of the suppression range.]{.underline}

## Exploratory Data Analysis

The graph below shows an unequal distribution of dispensing of medications of the 15 most widely distributed chemicals. Paracetamol is by far the most dispensed, with over double the amount of distributions compared to the asprin, the chemical in 4th place. The second and third most widely distributed medications are Atovastain and Omeprazole. Overall, the distribution has a long tail. Donepezil hydrochloride and Rivastigmine aren't highly distributed enough to make the top 15, with the former ranking 199th across the 5 years of data, and the latter ranking 475th over the same period.

```{r EDA1}
#| echo: false

# Bar chart of the most dispersed drugs
Data_ByChemical_count_ranked <- Data_ByChemical_count |>
  mutate(rank = row_number()) |>
  filter(rank <= 15 | Chemical %in% dementia_chemicals)

Data_ByChemical_count_ranked |>
  mutate(Chemical = fct_reorder(Chemical, NumDisps)) |>
  ggplot(aes(x = NumDisps, y = Chemical)) +
  geom_col() +
  geom_hline(
    yintercept = 2.5,
    colour = "red",
    linetype = "dotted",
    linewidth = 1
  ) +
  geom_text(
    aes(label = paste0("Rank ", rank)),
    hjust = -0.15,
    size = 3.5
  ) +
  scale_x_continuous(
    labels = label_comma()
  ) +
  expand_limits(
    x = max(Data_ByChemical_count_ranked$NumDisps) * 1.15
  ) +
  labs(
    title = "Bar chart of the most dispersed drugs",
    x = "Number of distributions",
    y = "Chemical",
    colour = NULL
  )
```

Looking at the line chart below showing rank over time, it can be seen that there are no changes to the rank of the top 3 medicines across the five years. Moving down the rankings, more variability is displayed. Colecalciferol rose from 7th to 4th, metoprobl siccinate moved down from 5th to 7th and Ibuprofin from 9th to 6th. This relates to two things:

- The long tailed nature of the ranked distribution means that the steeper decreases in medication top of the ranks are rank unphased relative to any changes over time.

- The similar distributions over time between subsequently ranked medications are more likely to have their ranks changed by upward or downward trends in distribution over time.

This suggests that any future modelling would benefit from having year of distribution as a regressor, but it would be less likely to change the rankings of the highest distributed drugs

```{r EDA2}
#| echo: false
#| fig-width: 10
#| fig-height: 6

# Rankings of the most frequently dispensed medications over time
Data_ByChemical_year_count |>
  ggplot(aes(
    x = YearDisp,
    y = Rank,
    colour = Chemical,
    group = Chemical
  )) +
  geom_line(
    linewidth = 1,
    alpha = 0.8
  ) +
  geom_point(size = 3) +
  scale_y_reverse(
    breaks = 1:15
  ) +
  labs(
    title = "Rankings of the most frequently dispensed medications over time",
    x = "Year",
    y = "Rank"
  )
```

The table below shows the distribution count and ranks of the two dementia medications of interest. There has been an upwards trend in distribution for both of these medications, while rivastigmine has had its rank reduced in recent years. This is because while the distribution of this medication has increased in these years, not as much as other medication that had a lower distribution count that rivastigmine in 2021.

```{r EDA3}
#| echo: false

Data_ByChemical_clean |>
  group_by(YearDisp, Chemical) |>
  summarise(NumDisps = sum(NumDisps)) |>
  ungroup() |>
  arrange(YearDisp, desc(NumDisps)) |>
  group_by(YearDisp) |>
  mutate(Rank = row_number()) |>
  filter(Chemical %in% dementia_chemicals) |>
  filter(Chemical %in% dementia_chemicals) |>
  pivot_wider(
    names_from = Chemical,
    values_from = c(NumDisps, Rank),
    names_glue = "{Chemical}_{.value}"
  ) |>
  rename(
    Year = YearDisp,
    `Donepezil hydrochloride distribution count` = `Donepezil hydrochloride_NumDisps`,
    `Rivastigmine distribution count` = `Rivastigmine_NumDisps`,
    `Donepezil hydrochloride distribution rank` = `Donepezil hydrochloride_Rank`,
    `Rivastigmine distribution rank` = `Rivastigmine_Rank`
  ) |>
  pander()
```

It is useful to know how frequently medications are distributed in a high concentration to a few health districts relative to an even spread across all health districts. One way to measure this is by using the proportion of distributions going toward the health district where most of the distributions occur. There are 19 health districts, and if the distributions occurred equally across said districts, this metric would take the value of 1/19. In the opposite extreme, if the medication was distributed in only 1 district, the metric would take the value of 1. Below is a histogram showing how each of the medications fall on this proportion scale.

```{r EDA4}
#| echo: false

# Distribution of the largest district dispensing shares
Chemical_concentration_region |>
  ggplot(aes(x = district_share)) +
  geom_histogram(
    bins = 30,
    colour = "white"
  ) +
  geom_vline(
    xintercept = median(
      Chemical_concentration_region$district_share
    ),
    linetype = "dashed"
  ) +
  annotate(
    "text",
    x = median(
      Chemical_concentration_region$district_share
    ),
    y = Inf,
    label = paste0("Median = ", round(
      median(
        Chemical_concentration_region$district_share
      ),
      digits = 4
    )),
    hjust = -0.1,
    vjust = 1.5
  ) +
  labs(
    title = "Distribution of the largest district dispensing shares",
    x = "Largest district share of dispensings",
    y = "Number of chemicals"
  )
```

The median value is 0.1695, meaning that a typical medication will have roughly 17% of their medication go towards the district that leads in distribution. The lowest observed proportion is `{r} Chemical_concentration_region |> pull(district_share) |> min()` meaning that no medication is equally distributed across all districts. There's a small group of medications that have a proportion value of 1, meaning that they are only distributed in a single district. Note that this metric doesn't speak to the number of distributions for these medications, and are likely very niche with an estimated mean of `{r} Chemical_concentration_region |> filter(district_share == 1) |> pull(NumDisps) |> mean() |> round(2)` and max of `{r} Chemical_concentration_region |> filter(district_share == 1) |> pull(NumDisps) |> max()` . The medications with these lower distributions are either very niche, available from retailers outside of pharmacies or not recorded in some districts.

Dementia as a risk is strongly associated with old age. This makes the districts with a higher count of aged population more relevant to this analysis than total population size in the health district. The graphs below show a positive relationship between aged population count and distributions of the two key dementia medications. This relationship is stronger for Donepezil Hydrochloride, but both show that there are some observations above the line of best fit on the higher end of the district aged population. These observations belong to the `{r}``dementia_population |> group_by(Chemical) |>` `slice_max(order_by = NumDisps, n = 5) |>` `pull(District) |> unique()` district, and as the years increase it can be seen diverging further away from the line of best fit. Overall, this high correlation means that the older population size accounts for a large amount of the inter-district variation.

```{r EDA5}
#| echo: false

# Population size against dispensing count
dementia_population |>
  ggplot(aes(
    x = population_65_plus,
    y = NumDisps,
    shape = factor(YearDisp),
    colour = factor(YearDisp)
  )) +
  geom_point() +
  geom_smooth(
    aes(group = 1),
    method = "lm",
    se = TRUE,
    colour = "black",
    fill = "grey",
    linetype = "dashed"
  ) +
  facet_wrap(
    ~ Chemical,
    scales = "free_y"
  ) +
  scale_x_continuous(
    labels = label_comma()
  ) +
  scale_y_continuous(
    labels = label_comma()
  ) +
  labs(
    title = "Dementia-medication dispensing and the population aged 65+",
    x = "District population aged 65 and older",
    y = "Number of dispensings",
    shape = "Year",
    colour = "Year"
  )
```

Crude rates were calculated as annual distributions per 1,000 people aged 65 years and over for the 2024 year. The table below shows the variation in these distribution rates across the different districts.

```{r EDA6}
#| echo: false

# 2024 Dispensing rates by district
dementia_population |>
  filter(YearDisp == 2024) |>
  ggplot(
    aes(
      x = dispensings_per_1000_65plus,
      y = factor(
        District,
        levels = rev(district_order)
      )
    )
  ) +
  geom_col(
    fill = "grey"
  ) +
  facet_wrap(
    ~ Chemical,
    scales = "free_x"
  ) +
  labs(
    title = "Dementia-medication dispensing rates by district in 2024",
    x = "Dispensings per 1,000 people aged 65+",
    y = NULL
  )
```

The first observation of note is that donepezil hydrochloride has a much higher rate than rivastigmine. Tairāwhiti has the highest overall rate when you take into account said larger rates for donepezil hydrochloride, as it is the district that has the highest rate for this medication. Canterbury and Hawke's Bay have the second and third highest rates respectively. Alternatively, Rivastigmine has the highest concentration of distribution in Canterbury, Taranaki and Southern. Taranaki doesn't stand out for being high or low with donepezil hydrochloride, and Tairāwhti has the lowest distribution rate for rivastigmine. This could be related to the high proportion of the population of Tairāwhti falling higher on the latest (2023) NZ deprivation scale as seen below.

```{r EDA7}
#| echo: false

# Convert deprivation back to long format for graphics
deprivation_long <- Deprivation_wide |>
  # Keep only deprivation categories 1–10
  filter(
    new_zealand_index_of_socioeconomic_deprivation %in%
      c(
        "1 - Least deprived",
        "2", "3", "4", "5", "6", "7", "8", "9",
        "10 - Most deprived"
      )
  ) |>
  # Convert the area columns into rows
  pivot_longer(
    cols = -new_zealand_index_of_socioeconomic_deprivation,
    names_to = "area",
    values_to = "count"
  ) |>
  group_by(area) |>
  mutate(
    proportion = count / sum(count)
  ) |>
  ungroup() |>
  # remove area instances that are more aggregated than health district
  filter(
    !area %in% c(
      "Total - New Zealand by health region/health district",
      "Central Region",
      "Northern Region",
      "Te Manawa Taki",
      "Te Waipounamu"
    )
  ) |>
  mutate(area = factor(area, levels = rev(district_order)))

# Heatmap of deprivation index across districts
deprivation_long |> 
  ggplot(
    aes(
      x = new_zealand_index_of_socioeconomic_deprivation,
      y = area,
      fill = proportion
    )
  ) +
  geom_tile(colour = "white") +
  scale_fill_viridis_c(
    labels = label_percent(accuracy = 1),
    name = "Population share"
  ) +
  labs(
    title = "Distribution of socioeconomic deprivation by area",
    x = "New Zealand Index of Socioeconomic Deprivation",
    y = NULL
  ) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

## Detailed Analysis Results

[Potential models:\
Count data suggests either Poisson model, or negative binomial if overdispersion needs to be accounted for (Create both and compare the diagnostics of the two)]{.underline}

[Inclusion of interactions. Why they were/weren't required]{.underline}

[Include confidence intervals for the different coefficients]{.underline}

[Limitations:]{.underline}

[Only 2023 census returns are informing the deprivation profiles for the health districts, whereas the pharmaceutical dispensings are annually temporal]{.underline}

[All data is summarized to the district level, so we can only look at trends in a region as opposed to any individual level information]{.underline}

## Conclusions and Recommendations
