---
title: "Pharmaceutical Report"
author: "Group 22"
format: typst
editor: visual
---

```{r setup}
#| include: false

# ---------------------------------------------------------------------------------
# Setup
# ---------------------------------------------------------------------------------

# Clear the environment
rm(list = ls())

# load the packages
pacman::p_load(
  tidyverse,           # Data wrangling + graphics
  tibble,              # Alternative to dataframes with slightly altered rules
  janitor,             # Tidy dataframe names
  scales,              # Graphics formatting
  MASS,                # Negative binomial models
  glmmTMB,             # Negative binomial mixed models
  pander               # clean tables
)

# Set the seed
set.seed(111)

# Suppress warnings
options(warn = -1)

# Load the data
Data_ByChemical <- read.csv("Data/Data_ByChemical.csv") |>
  as_tibble()
Deprivation <- read.csv("Data/Deprivation index 2023.csv") |>
  as_tibble() |>
  clean_names()
Retired_population <- read.csv("Data/Retired population annual estimates.csv") |>
  as_tibble() |>
  clean_names()


# Define deprivation order
deprivation_order <- c(
  "1 - Least deprived",
  "2", "3", "4", "5", "6", "7", "8", "9",
  "10 - Most deprived",
  "Not elsewhere included",
  "Total stated - New Zealand index of socioeconomic deprivation",
  "Total - New Zealand index of socioeconomic deprivation"
)

# Approximate geographic order from north to south
district_order <- c(
  "Northland",
  "Waitematā",
  "Auckland",
  "Counties Manukau",
  "Waikato",
  "Bay of Plenty",
  "Lakes",
  "Tairāwhiti",
  "Taranaki",
  "Whanganui",
  "MidCentral",
  "Hawke's Bay",
  "Capital, Coast and Hutt Valley",
  "Wairarapa",
  "Nelson Marlborough",
  "West Coast",
  "Canterbury",
  "South Canterbury",
  "Southern",
  "Area outside health region"
)

# Create a vector just for dementia medication
dementia_chemicals <- c(
  "Donepezil hydrochloride",
  "Rivastigmine"
)

# ---------------------------------------------------------------------------------
# Summaries
# ---------------------------------------------------------------------------------

# Remove the totals
Data_ByChemical_clean <- Data_ByChemical |>
  filter(District != "New Zealand") |>
  # Rename the districts for merging purposes
  mutate(
    District = case_when(
      District == "Tairawhiti" ~ "Tairāwhiti",
      District == "Waitemata" ~ "Waitematā",
      District %in% c("Capital & Coast", "Hutt Valley") ~ "Capital, Coast and Hutt Valley",
      .default = District
    ),
    NumDisps = as.numeric(
      case_when(
        NumDisps == "<6" ~ "3", 
        .default = NumDisps)
    ),
    NumPpl = as.numeric(
      case_when(
        NumPpl == "<6" ~ "3", 
        .default = NumPpl)
    )
  ) |>
  # Summarise the Capital, Coast and Hutt Valley instances
  group_by(Type, ChemID, Chemical, District, YearDisp, NHIComp) |>
  summarise(
    NumDisps = sum(NumDisps),
    NumPpl = sum(NumPpl)
  )

# ------------------------------------------------------------------------------
# Summaries
# ------------------------------------------------------------------------------

# 15 most common medicines by number of distributions
Common_chem_15 <- Data_ByChemical_clean |>
  group_by(Chemical) |>
  summarise(NumDisps = sum(NumDisps)) |>
  ungroup() |>
  slice_max(NumDisps, n = 15) |>
  pull(Chemical)

# - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
# Chemical dispension count
# - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 

Data_ByChemical_count <- Data_ByChemical_clean |>
  group_by(Chemical) |>
  summarise(NumDisps = sum(NumDisps)) |>
  ungroup() |>
  arrange(desc(NumDisps))

Data_ByChemical_year_count <- Data_ByChemical_clean |>
  group_by(YearDisp, Chemical) |>
  summarise(NumDisps = sum(NumDisps)) |>
  ungroup() |>
  arrange(YearDisp, desc(NumDisps)) |>
  group_by(YearDisp) |>
  mutate(Rank = row_number()) |>
  slice_min(Rank, n = 15) |>
  ungroup()

# - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
# Concentrated Medication by District
# - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 

# Find the regions where each medication has the highest concentration of
# distribution
Chemical_concentration_region <- Data_ByChemical_clean |>
  filter(District != "Unknown") |>
  group_by(Chemical, District) |>
  summarise(NumDisps = sum(NumDisps)) |>
  ungroup() |>
  group_by(Chemical) |>
  mutate(
    Chemical_total = sum(NumDisps),
    district_share = NumDisps / Chemical_total,
    districts_observed = n()
  ) |>
  slice_max(
    district_share, n = 1
  ) |>
  ungroup() |>
  arrange(desc(district_share))

# - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
# Deprivation pivot
# - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 

# Change the raw data to wide format
Deprivation_wide <- Deprivation |>
  dplyr::select(area, new_zealand_index_of_socioeconomic_deprivation, obs_value) |>
  pivot_wider(names_from = area, values_from = obs_value) |>
  mutate(
    new_zealand_index_of_socioeconomic_deprivation = factor(
      new_zealand_index_of_socioeconomic_deprivation,
      levels = deprivation_order,
      ordered = TRUE
    )
  ) |>
  arrange(new_zealand_index_of_socioeconomic_deprivation)

# Create a lookup of average deprivations per district
Deprivation_average <- Deprivation_wide |>
  mutate(
    index = parse_number(
      as.character(
        new_zealand_index_of_socioeconomic_deprivation
      )
    )
  ) |>
  filter(index %in% 1:10) |>
  pivot_longer(
    cols = -c(
      new_zealand_index_of_socioeconomic_deprivation,
      index
    ),
    names_to = "District",
    values_to = "population"
  ) |>
  group_by(District) |>
  summarise(
    average_deprivation = weighted.mean(
      index,
      population,
      na.rm = TRUE
    ),
    .groups = "drop"
  )

# - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
# Dementia medication
# - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 

# Create district/year population totals for people aged 65+
Population_65_plus <- Retired_population |>
  filter(
    sex == "Total people, sex",
    age %in% c(
      "65-69 years",
      "70-74 years",
      "75-79 years",
      "80-84 years",
      "85-89 years",
      "90 years and over"
    )
  ) |>
  mutate(
    # Make geographic names match the pharmaceutical data
    District = case_when(
      area == "Tairawhiti" ~ "Tairāwhiti",
      area == "Waitemata" ~ "Waitematā",
      area %in% c("Capital & Coast", "Hutt Valley") ~
        "Capital, Coast and Hutt Valley",
      area == "Area outside health district" ~
        "Area outside health region",
      .default = area
    )
  ) |>
  group_by(
    YearDisp = year_at_30_june,
    District
  ) |>
  summarise(
    population_65_plus = sum(obs_value)
  ) |>
  ungroup()

dementia_data <- Data_ByChemical_clean |>
  filter(
    Chemical %in% dementia_chemicals,
    District != "Unknown"
  ) |>
  group_by(YearDisp, District, Chemical) |>
  summarise(
    NumDisps = sum(NumDisps),
    NumPpl = sum(NumPpl)
  ) |>
  ungroup() |>
  left_join(
    Deprivation_average,
    by = "District"
  ) |>
  left_join(
    Population_65_plus,
    by = c("YearDisp", "District")
  )

# - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
# Population counts
# - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 

# Tidy the data
Retired_population_clean <- Retired_population |>
  filter(
    age != "65 years and over",
    area != "Area outside health district"
  ) |>
  mutate(
    YearDisp = year_at_30_june,
    age_group = age,
    age_lower = parse_number(age),
    population = obs_value,
    District = area
  ) |>
  dplyr::select(YearDisp, YearDisp, age_group, age_lower, District, population)

# Find the district populations for people aged 65 or over
population_65_plus <- Retired_population_clean |>
  group_by(
    YearDisp,
    District
  ) |>
  summarise(
    population_65_plus = sum(
      population
    )
  ) |>
  ungroup()

# Combine population and dispensing data
dementia_population <- dementia_data |>
  mutate(
    dispensings_per_1000_65plus =
      NumDisps / population_65_plus * 1000,
    
    people_per_1000_65plus =
      NumPpl / population_65_plus * 1000
  )
```

## Executive Summary

## Table of Contents

## Background and Data

[What is dementia and the importance of pharmaceutical treatement]{.underline}

[Importance of equity across socioeconomic status]{.underline}

[Pharmaceutical collection background]{.underline}

[Note that dispensing doesn't immediately imply that the medication was taken]{.underline}

[Association of dispensing between deprivation and time]{.underline}

[Background on the NZDep2023 population counts by deprivation decile and health disrtict, and the attributes included + supression rules and unknown districts]{.underline}

[The pharmaceutical data is a repeated cross sectional data frame showing how medicine was distributed in different regions over different years.]{.underline}

[Note that just because medicine was dispensed does not mean that it was taken.]{.underline}

[Additional district level variables to look into:\
Income, age, employment, education, racial profile (Probably not), family dependencies.]{.underline}

## Ethics, Privacy and Security

- [Ethical considerations:]{.underline}

  - [Any comparison of groups being done in this project shouldn't cause have any implications about financial status between said groups. For example, Māori people take less medication than Pākehā because they can't afford them.]{.underline}

  - [Given the nature of the data collection, no causal causal conclusions can be drawn. The analysis will only be able to create commentary about the trends and relationships between the variables.]{.underline}

- [Privacy considerations:]{.underline}

  - [We should not be removing the suppression layer from the count of medication distribution. While there is an oversight in how this data is provided that allows users to do this, and there is code demonstrating that this is possible in this project, the values were randomly adjusted so that for the example instance doesn't have the true count revealed.]{.underline}

  - [For all analysis, the suppressed values will be imputed. Currently the method of replacing these with the median of the suppression range.]{.underline}

## Exploratory Data Analysis

The graph below shows an unequal distribution of dispensing of medications of the 15 most widely distributed chemicals. Paracetamol is by far the most dispensed, with over double the amount of distributions compared to the asprin, the chemical in 4th place. The second and third most widely distributed medications are Atovastain and Omeprazole. Overall, the distribution has a long tail. Donepezil hydrochloride and Rivastigmine aren't highly distributed enough to make the top 15, with the former ranking 199th across the 5 years of data, and the latter ranking 475th over the same period.

```{r EDA1}
#| echo: false

# Bar chart of the most dispersed drugs
Data_ByChemical_count_ranked <- Data_ByChemical_count |>
  mutate(rank = row_number()) |>
  filter(rank <= 15 | Chemical %in% dementia_chemicals)

Data_ByChemical_count_ranked |>
  mutate(Chemical = fct_reorder(Chemical, NumDisps)) |>
  ggplot(aes(x = NumDisps, y = Chemical)) +
  geom_col() +
  geom_hline(
    yintercept = 2.5,
    colour = "red",
    linetype = "dotted",
    linewidth = 1
  ) +
  geom_text(
    aes(label = paste0("Rank ", rank)),
    hjust = -0.15,
    size = 3.5
  ) +
  scale_x_continuous(
    labels = label_comma()
  ) +
  expand_limits(
    x = max(Data_ByChemical_count_ranked$NumDisps) * 1.15
  ) +
  labs(
    title = "Bar chart of the most dispersed drugs",
    x = "Number of distributions",
    y = "Chemical",
    colour = NULL
  )
```

Looking at the line chart below showing rank over time, it can be seen that there are no changes to the rank of the top 3 medicines across the five years. Moving down the rankings, more variability is displayed. Colecalciferol rose from 7th to 4th, metoprobl siccinate moved down from 5th to 7th and Ibuprofin from 9th to 6th. This relates to two things:

- The long tailed nature of the ranked distribution means that the steeper decreases in medication top of the ranks are rank unphased relative to any changes over time.

- The similar distributions over time between subsequently ranked medications are more likely to have their ranks changed by upward or downward trends in distribution over time.

This suggests that any future modelling would benefit from having year of distribution as a regressor, but it would be less likely to change the rankings of the highest distributed drugs

```{r EDA2}
#| echo: false
#| fig-width: 10
#| fig-height: 6

# Rankings of the most frequently dispensed medications over time
Data_ByChemical_year_count |>
  ggplot(aes(
    x = YearDisp,
    y = Rank,
    colour = Chemical,
    group = Chemical
  )) +
  geom_line(
    linewidth = 1,
    alpha = 0.8
  ) +
  geom_point(size = 3) +
  scale_y_reverse(
    breaks = 1:15
  ) +
  labs(
    title = "Rankings of the most frequently dispensed medications over time",
    x = "Year",
    y = "Rank"
  )
```

The table below shows the distribution count and ranks of the two dementia medications of interest. There has been an upwards trend in distribution for both of these medications, while rivastigmine has had its rank reduced in recent years. This is because while the distribution of this medication has increased in these years, not as much as other medication that had a lower distribution count that rivastigmine in 2021.

```{r EDA3}
#| echo: false

Data_ByChemical_clean |>
  group_by(YearDisp, Chemical) |>
  summarise(NumDisps = sum(NumDisps)) |>
  ungroup() |>
  arrange(YearDisp, desc(NumDisps)) |>
  group_by(YearDisp) |>
  mutate(Rank = row_number()) |>
  filter(Chemical %in% dementia_chemicals) |>
  filter(Chemical %in% dementia_chemicals) |>
  pivot_wider(
    names_from = Chemical,
    values_from = c(NumDisps, Rank),
    names_glue = "{Chemical}_{.value}"
  ) |>
  rename(
    Year = YearDisp,
    `Donepezil hydrochloride distribution count` = `Donepezil hydrochloride_NumDisps`,
    `Rivastigmine distribution count` = `Rivastigmine_NumDisps`,
    `Donepezil hydrochloride distribution rank` = `Donepezil hydrochloride_Rank`,
    `Rivastigmine distribution rank` = `Rivastigmine_Rank`
  ) |>
  pander()
```

It is useful to know how frequently medications are distributed in a high concentration to a few health districts relative to an even spread across all health districts. One way to measure this is by using the proportion of distributions going toward the health district where most of the distributions occur. There are 19 health districts, and if the distributions occurred equally across said districts, this metric would take the value of 1/19. In the opposite extreme, if the medication was distributed in only 1 district, the metric would take the value of 1. Below is a histogram showing how each of the medications fall on this proportion scale.

```{r EDA4}
#| echo: false

# Distribution of the largest district dispensing shares
Chemical_concentration_region |>
  ggplot(aes(x = district_share)) +
  geom_histogram(
    bins = 30,
    colour = "white"
  ) +
  geom_vline(
    xintercept = median(
      Chemical_concentration_region$district_share
    ),
    linetype = "dashed"
  ) +
  annotate(
    "text",
    x = median(
      Chemical_concentration_region$district_share
    ),
    y = Inf,
    label = paste0("Median = ", round(
      median(
        Chemical_concentration_region$district_share
      ),
      digits = 4
    )),
    hjust = -0.1,
    vjust = 1.5
  ) +
  labs(
    title = "Distribution of the largest district dispensing shares",
    x = "Largest district share of dispensings",
    y = "Number of chemicals"
  )
```

The median value is 0.1695, meaning that a typical medication will have roughly 17% of their medication go towards the district that leads in distribution. The lowest observed proportion is `{r} Chemical_concentration_region |> pull(district_share) |> min()` meaning that no medication is equally distributed across all districts. There's a small group of medications that have a proportion value of 1, meaning that they are only distributed in a single district. Note that this metric doesn't speak to the number of distributions for these medications, and are likely very niche with an estimated mean of `{r} Chemical_concentration_region |> filter(district_share == 1) |> pull(NumDisps) |> mean() |> round(2)` and max of `{r} Chemical_concentration_region |> filter(district_share == 1) |> pull(NumDisps) |> max()` . The medications with these lower distributions are either very niche, available from retailers outside of pharmacies or not recorded in some districts.

Dementia as a risk is strongly associated with old age. This makes the districts with a higher count of aged population more relevant to this analysis than total population size in the health district. The graphs below show a positive relationship between aged population count and distributions of the two key dementia medications. This relationship is stronger for Donepezil Hydrochloride, but both show that there are some observations above the line of best fit on the higher end of the district aged population. These observations belong to the ``` {r}``dementia_population |> group_by(Chemical) |> ``` `slice_max(order_by = NumDisps, n = 5) |>` `pull(District) |> unique()` district, and as the years increase it can be seen diverging further away from the line of best fit. Overall, this high correlation means that the older population size accounts for a large amount of the inter-district variation.

```{r EDA5}
#| echo: false

# Population size against dispensing count
dementia_population |>
  ggplot(aes(
    x = population_65_plus,
    y = NumDisps,
    shape = factor(YearDisp),
    colour = factor(YearDisp)
  )) +
  geom_point() +
  geom_smooth(
    aes(group = 1),
    method = "lm",
    se = TRUE,
    colour = "black",
    fill = "grey",
    linetype = "dashed"
  ) +
  facet_wrap(
    ~ Chemical,
    scales = "free_y"
  ) +
  scale_x_continuous(
    labels = label_comma()
  ) +
  scale_y_continuous(
    labels = label_comma()
  ) +
  labs(
    title = "Dementia-medication dispensing and the population aged 65+",
    x = "District population aged 65 and older",
    y = "Number of dispensings",
    shape = "Year",
    colour = "Year"
  )
```

Crude rates were calculated as annual distributions per 1,000 people aged 65 years and over for the 2024 year. The table below shows the variation in these distribution rates across the different districts.

```{r EDA6}
#| echo: false

# 2024 Dispensing rates by district
dementia_population |>
  filter(YearDisp == 2024) |>
  ggplot(
    aes(
      x = dispensings_per_1000_65plus,
      y = factor(
        District,
        levels = rev(district_order)
      )
    )
  ) +
  geom_col(
    fill = "grey"
  ) +
  facet_wrap(
    ~ Chemical,
    scales = "free_x"
  ) +
  labs(
    title = "Dementia-medication dispensing rates by district in 2024",
    x = "Dispensings per 1,000 people aged 65+",
    y = NULL
  )
```

The first observation of note is that donepezil hydrochloride has a much higher rate than rivastigmine. Tairāwhiti has the highest overall rate when you take into account said larger rates for donepezil hydrochloride, as it is the district that has the highest rate for this medication. Canterbury and Hawke's Bay have the second and third highest rates respectively. Alternatively, Rivastigmine has the highest concentration of distribution in Canterbury, Taranaki and Southern. Taranaki doesn't stand out for being high or low with donepezil hydrochloride, and Tairāwhti has the lowest distribution rate for rivastigmine. This could be related to the high proportion of the population of Tairāwhti falling higher on the latest (2023) NZ deprivation scale as seen below.

```{r EDA7}
#| echo: false

# Convert deprivation back to long format for graphics
deprivation_long <- Deprivation_wide |>
  # Keep only deprivation categories 1–10
  filter(
    new_zealand_index_of_socioeconomic_deprivation %in%
      c(
        "1 - Least deprived",
        "2", "3", "4", "5", "6", "7", "8", "9",
        "10 - Most deprived"
      )
  ) |>
  # Convert the area columns into rows
  pivot_longer(
    cols = -new_zealand_index_of_socioeconomic_deprivation,
    names_to = "area",
    values_to = "count"
  ) |>
  group_by(area) |>
  mutate(
    proportion = count / sum(count)
  ) |>
  ungroup() |>
  # remove area instances that are more aggregated than health district
  filter(
    !area %in% c(
      "Total - New Zealand by health region/health district",
      "Central Region",
      "Northern Region",
      "Te Manawa Taki",
      "Te Waipounamu"
    )
  ) |>
  mutate(area = factor(area, levels = rev(district_order)))

# Heatmap of deprivation index across districts
deprivation_long |> 
  ggplot(
    aes(
      x = new_zealand_index_of_socioeconomic_deprivation,
      y = area,
      fill = proportion
    )
  ) +
  geom_tile(colour = "white") +
  scale_fill_viridis_c(
    labels = label_percent(accuracy = 1),
    name = "Population share"
  ) +
  labs(
    title = "Distribution of socioeconomic deprivation by area",
    x = "New Zealand Index of Socioeconomic Deprivation",
    y = NULL
  ) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

The plots below show how the average distribution rate per aged person changes for the two dementia medications over the deprivation index. For donepezil hydrochloride, the trend in the relationship between the distribution rate and the deprivation index increases as time increases, peaking at 2022, followed by a decreasing trend in subsequent years. Alternatively, these trends have consistently steeper downward trends as the years increase. This is a strong argument to test interaction effects between year and deprivation index when predicting the number of medication distributions.

```{r EDA8}
#| echo: false

# Scatterplot showing dispensing rate by average deprivation for each year and
# drug
dementia_population |>
  ggplot(
    aes(
      x = average_deprivation,
      y = dispensings_per_1000_65plus
    )
  ) +
  geom_point() +
  geom_smooth(
    method = "lm",
    se = TRUE,
    colour = "red",
    fill = "grey"
  ) +
  facet_grid(
    Chemical ~ YearDisp,
    scales = "free_y"
  ) +
  labs(
    title = "Deprivation and dementia medication dispensing rates",
    x = "Average deprivation",
    y = "Dispensings per 1,000 people aged 65+"
  )
```

These results remain exploratory. There are only about 19 districts per year, the same districts are observed repeatedly, and district deprivation changes little over the study period. Confidence intervals are wide, especially for rivastigmine, and the slopes may be sensitive to influential districts such as Tairāwhiti, Canterbury or Taranaki.

## Detailed Analysis Results

A variety of parametric models are tested to find which most appropriately fit the data. Creating the appropriate model allows for inference on the relationships between the different regressors. The table compares models that predict the number of dementia medication distributions. The Poisson models assume that the variance is equal to the mean, while the negative binomial models allow for additional variation. The mixed models also include a random intercept for district, which accounts for repeated observations and differences between districts that are not explained by the other variables. The models become more complex as interactions are added, allowing the relationships with year and deprivation to differ between medications. The final model also uses the population aged 65 years and over as an offset, allowing it to model dispensing rates rather than only raw counts.

```{r DA1}
#| echo: false

dementia_model_data <- dementia_data |>
  mutate(
    Year = YearDisp,
    # Centre deprivation so model intercepts are easier to interpret
    deprivation_c =
      average_deprivation - mean(average_deprivation),
    Chemical = factor(Chemical),
    District = factor(District),
    log_population_65_plus = log(.data$population_65_plus)
  )

# Model 0: differences between chemicals only
poisson_0 <- dementia_model_data |>
  glm(
    NumDisps ~ Chemical,
    family = poisson(link = "log"),
    data = _
  )

# Model 1: additive effects
poisson_1 <- dementia_model_data |>
  glm(
    NumDisps ~ Year + deprivation_c + Chemical,
    family = poisson(link = "log"),
    data = _
  )

# Model 2: deprivation effect can differ by year
poisson_2 <- dementia_model_data |>
  glm(
    NumDisps ~ Year * deprivation_c + Chemical,
    family = poisson(link = "log"),
    data = _
  )

# Model 3: time patterns can also differ by chemical
poisson_3 <- dementia_model_data |>
  glm(
    NumDisps ~
      Year * deprivation_c +
      Year * Chemical,
    family = poisson(link = "log"),
    data = _
  )

# Model 4: all two-way interactions
poisson_4 <- dementia_model_data |>
  glm(
    NumDisps ~
      Year * deprivation_c +
      Year * Chemical +
      deprivation_c * Chemical,
    family = poisson(link = "log"),
    data = _
  )

# Check Poisson overdispersion
check_poisson_dispersion <- function(model) {
  pearson_dispersion <- sum(residuals(model, type = "pearson")^2) / df.residual(model)
  
  tibble(
    residual_deviance = deviance(model),
    residual_df = df.residual(model),
    deviance_ratio = deviance(model) / df.residual(model),
    pearson_dispersion = pearson_dispersion
  )
}

check_poisson_dispersion(poisson_0)
check_poisson_dispersion(poisson_1)
check_poisson_dispersion(poisson_2)
check_poisson_dispersion(poisson_3)
check_poisson_dispersion(poisson_4)
# a Pearson dispersion of roughly 1 would indicate that the Poisson assumption
# of the mean equalling the variance is appropriate. All of these models have a
# Pearson dispersion well above 1000, indicating that there is substantial
# overdispersion for a Poisson model

# Additive negative binomial model
nb_1 <- dementia_model_data |>
  glm.nb(NumDisps ~ Year + deprivation_c + Chemical, data = _)

# Year-by-deprivation interaction
nb_2 <- dementia_model_data |>
  glm.nb(NumDisps ~ Year * deprivation_c + Chemical, data = _)

# Year trends can differ by chemical
nb_3 <- dementia_model_data |>
  glm.nb(NumDisps ~
           Year * deprivation_c +
           Year * Chemical, data = _)

# All substantively useful two-way interactions
nb_4 <- dementia_model_data |>
  glm.nb(NumDisps ~
           Year * deprivation_c +
           Year * Chemical +
           deprivation_c * Chemical,
         data = _)

# Poisson mixed model
mixed_poisson <- dementia_model_data |>
  glmmTMB(
    NumDisps ~
      Year * deprivation_c +
      Chemical +
      (1 | District),
    family = poisson(link = "log"),
    data = _
  )

# Negative binomial mixed model
mixed_nb_1 <- dementia_model_data |>
  glmmTMB(
    NumDisps ~
      Year * deprivation_c +
      Chemical +
      (1 | District),
    family = nbinom2(link = "log"),
    data = _
  )

# Allow chemical-specific time patterns
mixed_nb_2 <- dementia_model_data |>
  glmmTMB(
    NumDisps ~
      Year * deprivation_c +
      Year * Chemical +
      (1 | District),
    family = nbinom2(link = "log"),
    data = _
  )

# Include all relevant two-way interactions
mixed_nb_3 <- dementia_model_data |>
  glmmTMB(
    NumDisps ~
      Year * deprivation_c +
      Year * Chemical +
      deprivation_c * Chemical +
      (1 | District),
    family = nbinom2(link = "log"),
    data = _
  )

# Include aged population as an  offset
mixed_nb_4 <- dementia_model_data |>
  glmmTMB(
    NumDisps ~
      Year * deprivation_c +
      Year * Chemical +
      deprivation_c * Chemical +
      offset(log_population_65_plus) +
      (1 | District),
    family = nbinom2(link = "log"),
    data = _
  )

model_explanations <- tribble(
  ~model,           ~explanation,
  "poisson_0",      "Poisson model containing differences between the two medications only.",
  "poisson_1",      "Adds year and average deprivation as separate effects.",
  "poisson_2",      "Allows the relationship between deprivation and dispensing to change over time.",
  "poisson_3",      "Also allows the time trend to differ between medications.",
  "poisson_4",      "Also allows the deprivation relationship to differ between medications.",
  "nb_1",           "Negative binomial version of poisson_1, allowing for overdispersion.",
  "nb_2",           "Negative binomial version of poisson_2, allowing for overdispersion.",
  "nb_3",           "Negative binomial version of poisson_3, allowing for overdispersion.",
  "nb_4",           "Negative binomial version of poisson_4, allowing for overdispersion.",
  "mixed_poisson",  "Poisson model that also accounts for differences between districts.",
  "mixed_nb_1",     "Negative binomial mixed model with a year-by-deprivation interaction.",
  "mixed_nb_2",     "Also allows the time trend to differ between medications.",
  "mixed_nb_3",     "Also allows the deprivation relationship to differ between medications.",
  "mixed_nb_4",     "Adds the population aged 65 years and over as an offset, producing population-adjusted dispensing rates."
)

BIC(
  poisson_0,
  poisson_1,
  poisson_2,
  poisson_3,
  poisson_4,
  nb_1,
  nb_2,
  nb_3,
  nb_4,
  mixed_poisson,
  mixed_nb_1,
  mixed_nb_2,
  mixed_nb_3,
  mixed_nb_4
) |>
  as.data.frame() |>
  rownames_to_column("model") |>
  as_tibble() |>
  left_join(
    model_explanations,
    by = "model"
  ) |>
  arrange(BIC) |>
  relocate(explanation, .after = model) |>
  pander()
```

The best fitting model, based on having the lowest BIC, uses the number of distributions as the response. It treats the number of distributions as a negative binomial distributed variable that depends on several district-level regressors. These are year, average deprivation and medication. Interactions between these variables are also included. The log of the population aged 65 years or older is used as an offset, which accounts for differences in the size of the older population across districts. District is included as a random intercept because each district is observed multiple times.

The model suggests that the distribution rates for the two medications changed differently over time. At the average level of deprivation, the coefficient for year is 0.0293 for donepezil hydrochloride. After exponentiating this coefficient, the estimated rate ratio is approximately 1.030. This suggests that the distribution rate for donepezil hydrochloride increased by around 3.0% per year. The year coefficient for rivastigmine also includes the interaction between year and chemical. This gives a coefficient of 0.0293 - 0.0629 = -0.0336. The corresponding rate ratio is approximately 0.967, suggesting that the distribution rate for rivastigmine decreased by around 3.3% per year. This means that the two medications had different patterns over time.

The relationship between deprivation and distribution also differs between the medications. The deprivation coefficient cannot be interpreted directly because year has not been centred. The deprivation effect for a particular year can be found by combining the deprivation coefficient with the interaction between year and deprivation.

For donepezil hydrochloride in 2020, the estimated deprivation coefficient is approximately 0.245. This gives a rate ratio of approximately 1.28. Therefore, a one-unit increase in average district deprivation was associated with an estimated 28% increase in the donepezil hydrochloride distribution rate in 2020, after accounting for the other variables in the model. For rivastigmine, the interaction between deprivation and chemical also needs to be included. This gives an estimated deprivation coefficient of approximately -0.308 in 2020 and a rate ratio of approximately 0.735. Therefore, a one-unit increase in average district deprivation was associated with an estimated 26.5% decrease in the rivastigmine distribution rate. This supports the pattern seen in the exploratory analysis, where the deprivation relationships moved in opposite directions for the two medications.

The interaction between year and deprivation is negative. This suggests that the relationship between deprivation and donepezil hydrochloride distribution became less positive over time. It also means that the negative relationship between deprivation and rivastigmine distribution became stronger over time.

The random intercept for district has a standard deviation of 0.357. Exponentiating this gives approximately 1.43. This means that a district one standard deviation above the average district effect had an expected distribution rate around 43% higher, after accounting for the variables included in the model. There is therefore still variation between districts that is not explained by year, deprivation, medication or older population size.

The negative binomial dispersion parameter is 7.33. This allows the variance of the response to be larger than its mean. This is more suitable than a Poisson model because the earlier Poisson models showed substantial overdispersion (as was reflected in the poor BIC fit).

The large values of some coefficients, such as the intercept and deprivation coefficient, are mainly caused by using the calendar year directly. The intercept currently represents the expected value in year zero.

```{r DA2}
#| echo: false

(mixed_nb_4)
```

A limitation of this model is that while most of the attributes are available on an annual level, notably deprivation index is not. An improvement would be to apply the deprivation index on the year that's appropriate, but the data informing this deprivation metric isn't sourced annually. Instead, it is sourced from the NZ census which is run every 5 years.

Another limitation is that the data for distribution of medication is shared at the district level as opposed to the individual level. While a more granular view of the attributes would improve the accuracy of the model, and create stronger inferences, it becomes a violation of privacy if individuals medication status were available publicly on the individual level. Also to do with data availability, the model creates a relationship with the number of medication distributions with the regressors, as opposed to the medication taken by individuals wich may not be as relevant to interested parties.

## Conclusions and Recommendations
